An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-11-17 thesis/abstract.tex
updates

We study the impact of the matrix structure on
the cyclic reduction algorithm
and show that that precomputing
forward reduction coefficients can be
an especially effective strategy.
his is the abstract

>>>>

\emph{Compact} finite difference schemes are widely used
in the direct numerical simulation of fluid flows,
for their ability to better resolve the small scales of turbulence.
However, they can be expensive to evaluate and
difficult to parallelize.
In this work, we present an approach for the computation
of compact finite differences
on graphics processing units (GPUs).
We present a variant of the \emph{cyclic reduction} 
algorithm for solving the
tridiagonal linear systems that arise in such numerical schemes.
We study the impact of the matrix structure on
the cyclic reduction algorithm
and show that that precomputing
forward reduction coefficients can be
especially effective for obtaining good performance.
Our tridiagonal solver is able to outperform
the NVIDIA CUSPARSE
and the multithreaded Intel MKL tridiagonal solvers
on GPU and CPU respectively.
In addition,
we present a parallelization strategy for GPU-accelerated clusters,
and show scalabality
of a 3-D compact finite difference application
for up to 64 GPUs on the Palmetto cluster.

### 2015-11-17 thesis/acknowledgments.tex
updates

I acknowledge everyone.

>>>>

I owe my deepest thanks
to my advisor, Dr. Richard S. Miller,
for hist immense guidance and support;
and the members of my committee,
Dr. Gang Li and Dr. Lonny Thompson
for their valuable time.
I would like to acknowledge
Dr. Daniel Livescu at Los Alamos National Lab
for introducing us to this interesting problem,
for his helpful comments,
and for generous funding.
I also thank
Dr. Melissa Smith and Karan Sapra
for their helpful guidance,
and for organizing an excellent course on GPU computing.
I owe a great deal of gratitude to
Dr. Christopher Cox at Clemson University
for his guidance on a range of topics
and for many interesting conversations.
I would like to thank the
staff of the Cyberinfrastructure Technology Integration,
and especially
Dr. Galen Collier and Dr. Marcin Ziolkowski
for their valuable comments and support.
I owe thanks to Clemson University for the generous
allocation of compute time on the Palmetto cluster.
I acknowledge
Dr. K. N. Seetharamu and Dr. V. Krishna,
my advisors at PES Institute of Technology
for inspiring in me a passion for
scientific research.
I owe a special thanks to the
Software Carpentry Foundation
and its members,
for helping me be a
a more productive programmer,
a more deliberate learner,
a more effective teacher,
and a better person.

Finally,
I would like to thank my family and friends
for everything they have done for me.

### 2015-11-17 thesis/compact-finite-difference.tex
updates

Here, we discuss the application of the
tridiagonal solver developed in the previous section
to the evaluation of compact finite differences---which
are used widely in the direct numerical simulation
of fluid flows.

\begin{figure}
\begin{center}
\includegraphics[height=150pt]{img/computational-domain.eps}
\caption{Computational domain in 3-D.}
\label{fig:computational-domain}
\end{center}
\end{figure}
%
In typical DNS applications,
the computational domain considered
is a 3-dimensional,
regular, Cartesian grid with
$nx$, $ny$ and $nz$ grid points and grid spacing of
$dx$, $dy$ and $dz$ in the
respective coordinate directions,
\ref{fig:computational-domain},
Structured grids such as these
are widely used because they lend themselves naturally
to finite-difference methods.
To resolve all the spatial features of the flow,
the grid spacing is kept very small.
Consequently,
to simulate flows of practical sizes,
the number of grid points must be very large.
Because of the high computational cost
and memory requirements associated with
the large number of grid points
parallelism becomes mandatory.
We have spoken so far of GPUs as massively parallel systems.
However, each GPU has a very limited amount of global memory:
current NVIDIA Tesla accelerators have about 4 Gigabytes.
Thus, to accomodate the
larger problem sizes in DNS,
the use of \emph{multiple} GPUs is necessary,
which introduces a second level of parallelism.
In the next section,
we describe strategies to distribute
the problem domain among the GPUs
in such dual-level parallel systems.

@@

Both of these operations are amenable to parallel operations.
The right hand side calculation is a pointwise
\emph{stencil} operation, i.e.,
at every point in the domain,
the value of the right hand side
is computed as a combination of function values
at that point and its neighbouring points.
The stencil operations at individual points
are independent,
making the overall calculation highly parallelizable.
The solution of the tridiagonal systems,
is of course parallelizable
as has been discussed in previous sections.

@@

For a single GPU \ref{fig:compact-single-gpu},
when calculating the right hand sides,
each point in the grid is mapped to a single thread,
and each thread applies the required stencil operation
to compute the right hand side at that point.
We note that threads near the left and right boundaries
apply a different stencil from the interior threads.
The implementation of GPU kernels for
stencil operations such as these is a topic of wide study.
The most important considerations were brought out
in the paper by Micikevicius et al. ~\ref{micikevicius20093d}.
For solving the tridiagonal systems,
each thread block of the GPU is mapped to a
different \emph{grid line}
aligned along the
direction the derivatives are being calculated.
The right hand sides are stored
contiguously along these grid lines
(as in Fig. \ref{fig:block-mapping}),
and the modified cyclic reduction algorithm developed
can be used to solve for the derivatives.

>>>>

Here, we discuss the application of the
tridiagonal solver developed in the previous section
to the evaluation of compact finite differences---which
are used widely in the direct numerical simulation
of fluid flows.
%
\begin{figure}
\begin{center}
\includegraphics[height=150pt]{img/computational-domain.eps}
\caption{Computational domain in 3-D.}
\label{fig:computational-domain}
\end{center}
\end{figure}
%
In typical DNS applications,
the computational domain considered
is a 3-dimensional,
regular, Cartesian grid with
$nx$, $ny$ and $nz$ grid points and grid spacing of
$dx$, $dy$ and $dz$ in the
respective coordinate directions
(Fig. \ref{fig:computational-domain}).
Structured grids such as these
are widely used because they lend themselves naturally
to finite-difference methods.
To resolve all the spatial features of the flow,
the grid spacing is kept very small.
Consequently,
to simulate flows of practical sizes,
the number of grid points must be very large.
Because of the high computational cost
and memory requirements associated with
the large number of grid points
parallelism becomes mandatory.
We have spoken so far of GPUs as massively parallel systems.
However, each GPU has a very limited amount of global memory:
current NVIDIA Tesla K20 GPUs have about 4 Gigabytes.
Thus, to accomodate the
larger problem sizes in DNS,
the use of \emph{multiple} GPUs is necessary,
which introduces a second level of parallelism.
In the next section,
we describe strategies to distribute
the problem domain among the GPUs
in such dual-level parallel systems.

@@

Both of these operations are amenable to parallel operations.
The right hand side calculation is a pointwise
\emph{stencil} operation, i.e.,
at every point in the domain,
the value of the right hand side
is computed as a combination of function values
at that point and its neighbouring points.
The stencil operations at individual points
are independent,
making the overall calculation highly parallelizable.
The solution of the tridiagonal systems
is of course parallelizable,
as has been discussed in previous sections.

@@

For a single GPU \ref{fig:compact-single-gpu},
when calculating the right hand sides,
each point in the grid is mapped to a single thread,
and each thread applies the required stencil operation
to compute the right hand side at that point.
We note that threads near the left and right boundaries
apply a different stencil from the interior threads.
The implementation of GPU kernels for
stencil operations such as these is a topic of wide study.
The most important considerations were brought out
in the paper by Micikevicius et al. ~\cite{micikevicius20093d}.
For solving the tridiagonal systems,
each thread block of the GPU is mapped to a
different \emph{grid line}
aligned along the
direction the derivatives are being calculated.
The right hand sides are stored
contiguously along these grid lines
(as in Fig. \ref{fig:block-mapping}),
and the modified cyclic reduction algorithm developed
can be used to solve for the derivatives.

### 2015-11-17 thesis/introduction-and-background.tex
updates

As mentioned, DNS involves the use of extremely fine grids.
At each of the grid points,
various data about the flow must be stored---for
example, the geometric coordinates (x, y and z),
pressure, temperature and velocities at that point.
The amount of memory required to store this data
for even small problem sizes
can quickly exceed the capacity of modern workstations/PCs.
Thus, a \emph{distributed memory} parallel system
is generally required for performing DNS.
Here, the traditional approach has been
to distribute parallel tasks among individual CPU cores,
or groups of CPU cores
that share common memory spaces.
In the latter case,
each group of cores constitutesa \emph{shared memory} system,
and the overall system is referred to as a \emph{hybrid system}.

@@

    \item When the derivative is expressed as
        some combination of the function values
        on \emph{both} sides of $i$, i.e,
        at $\hdots$, $i-2$, $i-1$, $i$, $i-1$, $i-2$, $\hdots$,
        we refer to the approximation as a
        \emph{central} difference.
\end{enumerate}

The approximation of higher order derivatives generally requires
inclusion of a larger number of points in the
finite difference approximation
(referred to as the finite difference \emph{stencil}).
For a given order of derivative,
finite difference approximation of arbitrary stencil widths
may be derived,
with larger stencils associated with higher accuracy~\cite{fornberg1988generation}.

@@

Compact schemes express the derivative at a point $i$
in terms of
function values \emph{and derivatives}
at neighbouring points,
For example,
if $f_i$ represents the value of
a uniformly sampled function evaluated at the $i$th sample point,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:
%
\begin{equation}
\begin{split}
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
        f^{\prime}_i
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \hdots
    = 
    a\frac{f_{i+1} - f_{i-1}}{dx} + \
    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
    c\frac{f_{i+3} - f_{i-3}}{dx} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}
%
where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
are parameters that must satisfy certain constraints ~\cite{lele1992compact}.
We note that for equations of the form in Eq. \ref{eqn:general-compact},
the derivative at any points $i$ cannot be computed explicitly.
Instead, we must write similar equations for \emph{all} points $i$
in the range.
This results in a system of linear equations
with unknowns \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
which may be represented by the matrix system $A\bm{x}=\bm{d}$,
where $\bm{x}$ is the vector
\{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
$\bm{d}$ is a vector of right hand sides [Eq.\ref{eqn:general-compact}],
and $A$ is, in general, a \emph{banded} matrix.

@@

The parallel cyclic reduction (PCR) algorithm ~\cite{Zhang2010FTS}
has only the forward reduction phase.
The first forward reduction step is applied to
the odd and even indexed equations seperately,
yielding two reduced systems
of size $n/2$.
Forward reduction applied to both of these
systems then yields \emph{four} reduced
systems of size $n/4$.
The process is repeated till
$n/2$ 2-by-2 systems are left,
all of which can be solved trivially.
The PCR algorithm requires half the number of steps
required by CR ($log_2(n)$),
but does significantly more computation per-step.
Further, unlike CR,
PCR can be implemented free of bank conflicts.

>>>>

When solving for flows numerically,
at each of the grid points in the computational domain,
various data about the flow must be stored---for
example, the geometric coordinates (x, y and z),
pressure, temperature and velocities at that point.
For even small problem sizes,
the amount of memory required to store this data
can quickly exceed the capacity of modern workstations/PCs.
Thus, a \emph{distributed memory} parallel system
is generally required for performing DNS.
Here, the traditional approach has been
to distribute parallel tasks among individual CPU cores,
or groups of CPU cores
that share common memory spaces.
In the latter case,
each group of cores constitutes a \emph{shared memory} system,
and the overall system is referred to as a \emph{hybrid system}.

@@

    \item When the derivative is expressed as
        some combination of the function values
        on \emph{both} sides of $i$, i.e,
        at $\hdots$, $i-2$, $i-1$, $i$, $i-1$, $i-2$, $\hdots$,
        we refer to the approximation as a
        \emph{central} difference.
\end{enumerate}
%
The approximation of higher order derivatives generally requires
inclusion of a larger number of points in the
finite difference approximation
(referred to as the finite difference \emph{stencil}).
For a given order of derivative,
finite difference approximation of arbitrary stencil widths
may be derived,
with larger stencils associated with higher accuracy~\cite{fornberg1988generation}.
In the evaluation of spatial derivatives,
the above schemes are referred to as \emph{explicit}
schemes, as the derivative at each point
can be expressed as some explicit combination of function values.

Compact schemes express the derivative at a point $i$
in terms of
function values \emph{and derivatives}
at neighbouring points,
i.e., the derivative is expressed \emph{implicitly}.
For example,
if $f_i$ represents the value of
a uniformly sampled function evaluated at the $i$th sample point,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:
%
\begin{equation}
\begin{split}
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
        f^{\prime}_i
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \hdots
    = 
    a\frac{f_{i+1} - f_{i-1}}{dx} + \
    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
    c\frac{f_{i+3} - f_{i-3}}{dx} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}
%
where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
are parameters that must satisfy certain constraints ~\cite{lele1992compact}.
We note that for equations of the form in Eq. \ref{eqn:general-compact},
the derivative at any points $i$ cannot be computed explicitly.
Instead, we must write similar equations for \emph{all} points $i$
in the range.
This results in a system of linear equations
with unknowns \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
which may be represented by the matrix system $A\bm{x}=\bm{d}$,
where $\bm{x}$ is the vector
\{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
$\bm{d}$ is a vector of right hand sides [Eq.\ref{eqn:general-compact}],
and $A$ is, in general, a \emph{banded} matrix.

@@

@@

The parallel cyclic reduction (PCR) algorithm
has only the forward reduction phase.
The first forward reduction step is applied to
the odd and even indexed equations seperately,
yielding two reduced systems
of size $n/2$.
Forward reduction applied to both of these
systems then yields \emph{four} reduced
systems of size $n/4$.
The process is repeated till
$n/2$ 2-by-2 systems are left,
all of which can be solved trivially.
The PCR algorithm requires half the number of steps
required by CR ($log_2(n)$),
but does significantly more computation per-step.
Further, unlike CR,
PCR can be implemented free of bank conflicts
~\cite{Zhang2010FTS}.

### 2015-11-17 thesis/proposed-tridiagonal-algorithm.tex
updates

\begin{equation} \label{eqn:toeplitz-matrix}
A = 
\begin{pmatrix}
     b_1 & c_1  \\
     a_0 & b_0  &  c_0  \\
         & a_0  &  b_0 &  c_0  \\
         &      &  a_0 &  b_0 &  c    \\
         &      &      &      &  \ddots \\
         &      &      &      &     &  \ddots  \\
         &      &      &      &     &  a_n  &  b_n
\end{pmatrix}
\end{equation}
%
where, in general
\begin{align*}
    a_0 &\neq a_n  \\
    b_1 &\neq b_0 \neq b_n \\
    c_1 &\neq c_0 
\end{align*}
%
We refer to matrices with this specific structure as
\emph{near-Toeplitz tridiagonal matrices},
and the corresponding linear systems as
\emph{near-Toeplitz tridiagonal systems}.
These matrices appear in a wide range of applications
~\cite{sun1995application}
such as alternating direction implicit methods,
line relaxation methods,
and numerical solutions to one-dimensional differential equations.
In this chapter,
we will describe an efficient solver for solving such
near-Toeplitz tridiagonal systems efficiently on GPUs.

\subsection{Forward reduction}

@@

@@

This implementation works entirely on the GPU's global memory,
Here, we define two kernels - one for the forward reduction step,
and the other for the backward substitution step.
Each kernel is called $log_2(n)-1$ times.
An extra call to the
forward reduction kernel performs
the two-by-two solve.
At each step,
the size of the thread blocks is determined by the \emph{stride}
between elements accessed at that step.
For the forward reduction phase,
we use $n/2$ threads per block for the first step,
$n/4$ threads for the second step, and so on.
The pattern is reversed for the backward substitution phase,
beginning with 2 threads per block for the first step.
Although this ensures that there are no inactive threads
at any stage,
the occupancy of the GPU is still very low
during the end of forward reduction and the end of backward substitution.

\begin{listing}[H]
\begin{minted}[fontsize=\footnotesize]{python}
stride = 1
for i in range(n-1):
    stride *= 2
    forwardReductionKernel(num_blocks=nrhs, block_size=n/stride)

solveTwoByTwo()

for i in range(n-1):
    stride /= 2
    backSubstitutionKernel(num_blocks=nrhs, block=n/stride)
\end{minted}
\end{listing}

The precomputed coefficient arrays and right hand
are accessed by the kernels from global memory.
The kernel suffers from
strided memory access for the right-hand side,
but the precomputed coefficient values are accessed
without major coalescing problems.
Further, by precomputing the forward reduction coefficients,
we greatly reduce the number of computations
(and thus the number of uncoalesced memory accesses).

>>>>

\begin{equation} \label{eqn:toeplitz-matrix}
A = 
\begin{pmatrix}
     b_1 & c_1  \\
     a_0 & b_0  &  c_0  \\
         & a_0  &  b_0 &  c_0  \\
         &      &  a_0 &  b_0 &  c    \\
         &      &      &      &  \ddots \\
         &      &      &      &     &  \ddots  \\
         &      &      &      &     &  a_n  &  b_n
\end{pmatrix}
\end{equation}
%
where, in general
\begin{align*}
    a_0 &\neq a_n  \\
    b_1 &\neq b_0 \neq b_n \\
    c_1 &\neq c_0 
\end{align*}
%
We refer to matrices with this specific structure as
\emph{near-Toeplitz tridiagonal matrices},
and the corresponding linear systems as
\emph{near-Toeplitz tridiagonal systems}.
These matrices appear in a wide range of applications
~\cite{sun1995application}
such as alternating direction implicit methods,
line relaxation methods,
and numerical solutions to one-dimensional differential equations.
We describe an efficient solver for solving such
near-Toeplitz tridiagonal systems efficiently on GPUs.
\subsection{Forward reduction}

@@

In this section, we describe the implementation of a GPU solver
for solving a given near-Toeplitz tridiagonal system
for multiple right hand sides.
Our implementation uses
NVIDIA CUDA platform for programming the GPU,
but is easily translated to OpenCL.
The Python programming language is used
to interface with CUDA, by use of the
PyCUDA \cite{kloeckner_pycuda_2012} library.
We develop two approaches based on the common idea
of precomputed forward reduction coefficients---one that leverages
the GPU's \emph{shared memory},
and the other working directly with global memory.
The GPU kernels for both implementations are relatively
straightforward and compact,
spanning no more than 100 lines of code.

@@

This implementation works entirely on the GPU's global memory,
Here, we define two kernels - one for the forward reduction step,
and the other for the backward substitution step.
Each kernel is called $log_2(n)-1$ times.
An extra call to the
forward reduction kernel performs
the two-by-two solve.
At each step,
the size of the thread blocks is determined by the \emph{stride}
between elements accessed at that step.
For the forward reduction phase,
we use $n/2$ threads per block for the first step,
$n/4$ threads for the second step, and so on.
The pattern is reversed for the backward substitution phase,
beginning with 2 threads per block for the first step.
Although this ensures that there are no inactive threads
at any stage,
the occupancy of the GPU is still very low
during the end of forward reduction and the end of backward substitution.
The precomputed coefficient arrays and right hand
are accessed by the kernels from global memory.
The kernel suffers from
strided memory access for the right-hand side,
but the precomputed coefficient values are accessed
without major coalescing problems.
Further, by precomputing the forward reduction coefficients,
we greatly reduce the number of computations
(and thus the number of uncoalesced memory accesses).

### 2015-11-17 thesis/results.tex
updates

\section{Validation}
    \subsection{Validation of tridiagonal solver}
    \subsection{Validation of compact finite difference application}
        \subsubsection{Tests for individual kernels}
        \subsubsection{Tests for communication routines}
        \subsubsection{Test functions for derivatives}
        \subsubsection{Order of convergence tests}
    \subsubsection{Validation results}

\section{Performance of GPU tridiagonal solver}
\section{Performance of compact finite difference application}

>>>>

\section{Performance of GPU tridiagonal solver}

\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/bench-2d.eps}
\caption{Relative solver performance for 2-D problems. Relative time defined as:}
$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
\label{fig:bench-2d}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/bench-3d.eps}
\caption{Relative solver performance for 3-D problems. Relative time defined as:}
$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
\label{fig:bench-3d}
\end{center}
\end{figure}

In this section we present a performance overview
of our Near-Toeplitz tridiagonal solver (NEATO)
against a multi-threaded Intel MKL solver and
the CUSPARSE GPU solver
(\texttt{dgtsv} and \texttt{dgtsvStridedBatch} respectively).
The MKL solver is based on Gaussian Elimination with partial pivoting,
Boundary conditions may prevent
the matrix from being symmetric and/or diagonally dominant,
precluding the use of more specialized tridiagonal solvers.
and the CUSPARSE solver uses a combination of
Cyclic Reduction and Parallel Cyclic Reduction.
Both solvers are optimized for the underlying architectures.

The CPU code is compiled with the Intel C compiler (version 15.0),
and run (with OpenMP support) on up to
16 independent cores of the the same shared-memory node
(one thread per core).
The CPU is an
Intel Xeon Processor E5-2670 v2 (2.50 GHz, 25 MB Smart Cache).
GPU code is compiled with the CUDA toolkit (version 6.5.14),
and run on the
NVIDIA Tesla K20 Accelerator.
When measuring GPU performance (both CUSPARSE and NEATO),
we do \emph{not} include the cost
of data transfer between the CPU and GPU.
\texttt{-O2} level compiler optimizations are turned on for both
CPU and GPU code;
no further optimization options are enabled in either case.
Of course, we use double precision for all solvers.
The timings reported are
kernel \emph{execution} times, i.e.,
the time for all kernel(s) to execute completely
before returning the program control to the CPU.
All timings are averaged over 100 tridiagonal solves.

\subsection{NEATO: global memory v/s shared memory performance}

\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/global-vs-shared-2d.eps}
\caption{Comparison of global memory and shared
    memory implementations of NEATO (2D problems).}
\label{fig:global-vs-shared-2d}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/global-vs-shared-3d.eps}
\caption{Comparison of global memory and shared memory
    implementations of NEATO (3D problems).}
\label{fig:global-vs-shared-3d}
\end{center}
\end{figure}

In Figs. \ref{fig:global-vs-shared-2d} and \ref{fig:global-vs-shared-3d},
we report the performance of the two solvers for the case
$N_{rhs} = n$ and $N_{rhs} = n^2$.
These cases correspond to tridiagonal systems
arising in 2-D and 3-D problems respectively.
We note that the shared memory implementation
offers better performance in nearly all cases.
However, the relative speedup from using
shared memory diminishes with increasing problem size.
For larger problem sizes,
the synchronization costs associated with inactive threads
leads to poor shared memory performance.

\subsection{Comparison of NEATO with Intel MKL and CUSPARSE solvers}

In Fig. \ref{fig:bench-2d} and \ref{fig:bench-3d},
we provide the relative performance of
Intel MKL and CUSPARSE solvers and compare against
the NEATO shared memory implementation.
The relative performance for
each problem size is obtained by
normalizing the solver timings
by the timing for the fastest solver for that problem size.
Table \ref{table:bench} shows the timings of the various
solvers to solve different problem sizes.

\begin{sidewaystable}[]
\scriptsize
\centering
\caption{Performance of Intel MKL, CUSPARSE and NEATO solvers.}
\label{table:bench}
\input{table/bench.tex}
\end{sidewaystable}

\section{Performance of compact finite difference application}

The experiments were run on the
Clemson University Palmetto Cluster,
using NVIDIA Tesla K20 and K40 GPUs.
The K40 GPUs were used for the largest problem sizes.
Each compute node on the cluster is equipped with up to 2 GPUs,
and nodes are connected by 56 Gbps Infiniband interconnect.
We use Open MPI 1.8.1 configured with OpenFabrics support.

The timings reported are \emph{wall clock} times
with global synchronization between processes performed 
before and after evaluation of the derivatives.
All timings are averaged over 100 evaluations of the function derivatives
in each coordinate direction.
We make it clear that our reported
problem sizes represent the \emph{actual size of problem data}.
Although it may be considered sufficient to
run tests on a single line of processes
for measuring the compact finite difference solver performance,
we set up and solve the problem for the entire computational domain.
In the context of a larger simulation,
global synchronization between the processes
is typically performed before and after the evalaution of derivatives,
and it is important to consider the related overhead.

Figures \ref{fig:compact-profiling-1024-64}-\ref{fig:compact-profiling-2048-64}
show the time taken by the different steps of
the compact finite difference solver.
We note that for the larger problem size,
the evaluation of the primary systems (using the NEATO solver)
constitutes a larger majority of the total runtime,
which justifies our efforts in optimizing the tridiagonal solver.
For evaluation of the derivatives in the y- and z- directions,
we note that a significant portion of the runtime is dedicated
to performing permutations of the input and output data.
We attribute this to our naive implementation of the
permutation kernels (no shared memory).
Figures \ref{fig:strong-scaling-256}-\ref{fig:weak-scaling-256}
show the strong and weak scaling of the compact finite difference solver
for evaluating derivatives in all three coordinate directions.
We note good scaling for both 2-D and 3-D cases.
The strong scaling for 3-D problems is somewhat better
as the GPU is kept busy for all problem sizes.

\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/profiling-1024-64.eps}
\caption{Solving problem sized $1024^3$ on 64 GPUs}
\label{fig:compact-profiling-1024-64}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/profiling-2048-64.eps}
\caption{Solving problem sized $2048^3$ on 64 GPUs}
\label{fig:compact-profiling-2048-64}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/strong-scaling-256.eps}
\caption{Strong scaling for multi-GPU compact finite difference, problem size: $256^3$.}
\label{fig:strong-scaling-256}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/strong-scaling-512.eps}
\caption{Strong scaling for multi-GPU compact finite difference, problem size: $512^3$.}
\label{fig:strong-scaling-512}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/weak-scaling-128.eps}
\caption{Weak scaling for multi-GPU compact finite difference, problem size: $128^3$ per process.}
\label{fig:weak-scaling-128}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/weak-scaling-256.eps}
\caption{Weak scaling for multi-GPU compact finite difference, problem size: $256^3$ per process.}
\label{fig:weak-scaling-256}
\end{center}
\end{figure}
